---
title: " Simulated Reflectance"
author: "Tina Lasisi"
date: today
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# Load required libraries
library(tidyverse)
library(knitr)
library(scales)
library(patchwork)
library(ggbeeswarm)
library(ggridges)
library(ggmulti)
library(paletteer)
library(stringr)

# Get available palettes for potential use
palettedf <- tibble(paletteer::palettes_d_names)

# Set global chunk options for the document
knitr::opts_chunk$set(
  echo = FALSE,        # Don't show code in output
  include = TRUE,      # Include output
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Don't show warnings
  message = FALSE,     # Don't show messages
  fig.retina = 2,      # Higher resolution figures
  fig.width = 8,       # Default figure width
  fig.height = 6,      # Default figure height
  out.width = "100%"   # Full width in HTML output
)
```

```{r functions}
# ==============================================================================
# UTILITY FUNCTIONS
# ==============================================================================

#' Save plot with explicit filename
#' 
#' @param plot The ggplot object to save
#' @param plot_path Directory to save the plot in
#' @param filename Name of the file to save
#' @param width Width of the saved plot in inches
#' @param height Height of the saved plot in inches
pltsave_func <- function(plot, plot_path, filename, width, height){
  ggsave(
    filename = filename,
    plot = plot, 
    path = plot_path,
    width = width,
    height = height
  )
}

# ==============================================================================
# DATA PROCESSING FUNCTIONS
# ==============================================================================

#' Process raw spectral data from CSV
#' 
#' This function reads a CSV file of spectral data and:
#' 1. Renames columns to wavelength values (W380, W382, etc.)
#' 2. Converts values to percentages (multiply by 100)
#' 
#' @param file_name Path to the CSV file
#' @return A data frame with renamed columns and percentage values
processData <- function(file_name) {
  # Read the CSV file (all columns as numeric)
  data <- read.csv(file_name, colClasses = "numeric")
  
  # Generate wavelength column names (W380, W382, W384, etc.)
  # Starting at 380nm with 2nm intervals
  new_names <- paste0("W", seq(from = 380, 
                               to = (380 + (2 * (ncol(data) - 1))), 
                               by = 2))
  
  # Rename columns and convert to percentages
  processed_data <- data %>%
    set_names(new_names) %>% 
    mutate_all(~ .x * 100)
  
  return(processed_data)
}

#' Calculate melanin index and prepare data for plotting
#' 
#' This function:
#' 1. Calculates the melanin index (CMI) using wavelengths 640-670nm
#' 2. Filters data to reasonable CMI range (20-130)
#' 3. Samples data to prevent plot overload
#' 4. Converts to long format for ggplot
#' 
#' @param data Processed spectral data frame
#' @param sample_size Number of spectra to sample (default 400)
#' @return Long-format data frame with id, wavelength, reflectance, and cmi
calculateMIndex <- function(data, sample_size = 400) {
  # Calculate melanin index using the formula from the literature
  # This uses the red region of the spectrum (640-670nm)
  data_processed <- data %>% 
    mutate(
      # Average reflectance in the red region
      eqn1 = ((W650 + W660 + (0.5 * W640) + (0.5 * W660)) / 3) / 100,
      # Convert to melanin index
      cmi = 100 * log10(1 / eqn1)
    )
  
  # Prepare data for plotting
  data_long <- data_processed %>%
    mutate(id = row_number()) %>%
    # Filter to reasonable melanin index range
    filter(cmi >= 20 & cmi <= 130) %>% 
    # Sample to prevent plot overload
    slice_sample(n = sample_size) %>% 
    # Convert to long format
    pivot_longer(starts_with("W"), 
                 names_to = "wavelength", 
                 values_to = "reflectance") %>% 
    # Extract numeric wavelength values
    mutate(wavelength = as.numeric(str_remove(wavelength, "\\D+"))) %>% 
    # Focus on visible spectrum
    filter(wavelength >= 400 & wavelength <= 700) %>%
    select(id, everything())
  
  return(data_long)
}

# ==============================================================================
# PLOTTING FUNCTIONS
# ==============================================================================

#' Generate skin spectra plot
#' 
#' Creates a plot showing reflectance spectra colored by melanin index
#' 
#' @param data Long-format data with wavelength, reflectance, and cmi
#' @return A ggplot object
generateSkinSpectraPlot <- function(data) {
  plot <- ggplot(data, aes(x = wavelength, y = reflectance, 
                           group = id, color = cmi)) +
    geom_line(alpha = 0.6) +  # Add transparency to handle overlapping lines
    # Color gradient from dark to light skin
    scale_color_gradientn(
      colours = c("#42230c", "#a15c33", "#fbf7ec"),
      values = rescale(x = c(50, 40, 20)),
      oob = squish,
      limits = c(20, 130)
    ) +
    expand_limits(y = 0) +
    theme_classic() +
    labs(
      x = "Wavelength (nm)",
      y = "Percent Reflectance of Skin",
      color = "Melanin Index"
    ) +
    guides(color = guide_colorbar(reverse = TRUE)) +
    # Highlight the region used for melanin index calculation
    annotate("rect", xmin = 640, xmax = 670, ymin = 0, ymax = 70,
             alpha = 0.3, fill = "#ff0000") +
    labs(
      title = "Spectra from subsample of 400 simulated skin tones",
      subtitle = "Red region (640-670nm) is used to calculate melanin index"
    )
  
  return(plot)
}

#' Wrapper function to process file and generate plot
#' 
#' @param file_name Path to the CSV file
#' @return A ggplot object
generateSkinSpectraPlotWrapper <- function(file_name) {
  # Process the data
  processed_data <- processData(file_name)
  
  # Calculate melanin index
  data_with_mindex <- calculateMIndex(processed_data)
  
  # Generate the plot
  plot <- generateSkinSpectraPlot(data_with_mindex)
  
  return(plot)
}

# ==============================================================================
# FUNCTION TO CREATE ALL THREE PLOTS
# ==============================================================================

#' Create all three dataset plots for side-by-side comparison
#' 
#' @param folder_path Base directory containing the data folders
#' @return A list of three plots
createThreeDatasetPlots <- function(folder_path) {
  
  plots <- list()
  
  # Find and process the three specific files
  train_file <- list.files(file.path(folder_path, "train"), 
                          pattern = "dataset.*\\[0\\.000-1\\.000\\].*\\[0\\.650-0\\.800\\].*\\.csv$", 
                          full.names = TRUE)[1]
  
  test_wide_file <- list.files(file.path(folder_path, "test"), 
                               pattern = "dataset.*\\[0\\.000-1\\.000\\].*\\[0\\.650-0\\.800\\].*\\.csv$", 
                               full.names = TRUE)[1]
  
  test_narrow_file <- list.files(file.path(folder_path, "test"), 
                                pattern = "dataset.*\\[0\\.001-0\\.035\\].*\\[0\\.700-0\\.780\\].*\\.csv$", 
                                full.names = TRUE)[1]
  
  # Process each file
  if (!is.na(train_file)) {
    processed_data <- processData(train_file)
    data_with_mindex <- calculateMIndex(processed_data)
    plot1 <- generateSkinSpectraPlot(data_with_mindex) +
      labs(title = "TRAINING Dataset",
           subtitle = "Wide ranges: Thickness 0-1.0, Eumelanin 0.65-0.80") +
      theme(plot.title = element_text(size = 12, face = "bold"),
            plot.subtitle = element_text(size = 9),
            legend.position = "bottom")
    plots$train <- plot1
  }
  
  if (!is.na(test_wide_file)) {
    processed_data <- processData(test_wide_file)
    data_with_mindex <- calculateMIndex(processed_data)
    plot2 <- generateSkinSpectraPlot(data_with_mindex) +
      labs(title = "TEST Dataset (Wide)",
           subtitle = "Same as training: Thickness 0-1.0, Eumelanin 0.65-0.80") +
      theme(plot.title = element_text(size = 12, face = "bold"),
            plot.subtitle = element_text(size = 9),
            legend.position = "bottom")
    plots$test_wide <- plot2
  }
  
  if (!is.na(test_narrow_file)) {
    processed_data <- processData(test_narrow_file)
    data_with_mindex <- calculateMIndex(processed_data)
    plot3 <- generateSkinSpectraPlot(data_with_mindex) +
      labs(title = "TEST Dataset (Narrow)",
           subtitle = "Narrow TH & EU ranges: Thickness 0.001-0.035, Eumelanin 0.70-0.78") +
      theme(plot.title = element_text(size = 12, face = "bold"),
            plot.subtitle = element_text(size = 9),
            legend.position = "bottom")
    plots$test_narrow <- plot3
  }
  
  return(plots)
}

#' Create summary statistics plot instead of individual lines
#' 
#' This function creates a ribbon plot showing mean ± SD to handle large datasets
#' 
#' @param data Long-format data with wavelength, reflectance, and cmi
#' @return A ggplot object with summary statistics
generateSummarySpectraPlot <- function(data) {
  # Calculate summary statistics by wavelength
  summary_data <- data %>%
    group_by(wavelength) %>%
    summarise(
      mean_reflectance = mean(reflectance),
      sd_reflectance = sd(reflectance),
      min_reflectance = min(reflectance),
      max_reflectance = max(reflectance),
      q25 = quantile(reflectance, 0.25),
      q75 = quantile(reflectance, 0.75),
      .groups = 'drop'
    )
  
  # Create plot with ribbon showing variability
  plot <- ggplot(summary_data, aes(x = wavelength)) +
    # Show full range as light ribbon
    geom_ribbon(aes(ymin = min_reflectance, ymax = max_reflectance), 
                alpha = 0.2, fill = "gray70") +
    # Show interquartile range as darker ribbon
    geom_ribbon(aes(ymin = q25, ymax = q75), 
                alpha = 0.3, fill = "gray50") +
    # Show mean ± 1 SD
    geom_ribbon(aes(ymin = mean_reflectance - sd_reflectance, 
                    ymax = mean_reflectance + sd_reflectance), 
                alpha = 0.4, fill = "steelblue") +
    # Show mean line
    geom_line(aes(y = mean_reflectance), size = 1.2, color = "darkblue") +
    # Highlight melanin index region
    annotate("rect", xmin = 640, xmax = 670, ymin = 0, ymax = 70,
             alpha = 0.3, fill = "#ff0000") +
    theme_classic() +
    labs(
      x = "Wavelength (nm)",
      y = "Percent Reflectance of Skin",
      title = "Summary of skin reflectance spectra",
      subtitle = "Mean (dark line) ± SD (blue), IQR (dark gray), and full range (light gray)"
    )
  
  return(plot)
}
```

# Overview of Simulated Skin Reflectance Datasets

This analysis examines three distinct datasets of simulated skin reflectance spectra:

1. **Training Dataset**: Wide parameter ranges for model training
2. **Test Dataset (Wide)**: Same parameter ranges as training for validation
3. **Test Dataset (Narrow)**: Restricted parameter ranges to test model performance on a specific subset

**Note**: All datasets measure reflectance from 380-1000nm wavelength range, though plots focus on the visible spectrum (400-700nm).

**Terminology**: "Parameter range" refers to the minimum and maximum values allowed for each skin property in the simulation. For example, thickness range 0.000-1.000 means skin thickness can vary anywhere from 0 to 1 (in appropriate units).

## Dataset Parameters

### Parameters that remain CONSTANT across all datasets:
- **Melanin (ME)**: 0.001-0.430 - Total melanin content controlling overall skin darkness
- **Blood Volume (BL)**: 0.000-0.200 - Amount of blood in skin affecting redness/pinkness
- **Oxygenation (OX)**: 0.900-1.000 - Blood oxygen saturation level (high = red, low = blue)

### Parameters that VARY between datasets:
- **Thickness (TH)**: Physical thickness of skin layer affecting how light scatters
  - Wide range: 0.000-1.000 (full physiological range)
  - Narrow range: 0.001-0.035 (very thin skin only)
- **Eumelanin (EU)**: Fraction of melanin that is eumelanin (brown/black) vs pheomelanin (red/yellow)
  - Wide range: 0.650-0.800 (moderate to high eumelanin)
  - Narrow range: 0.700-0.780 (high eumelanin only)

## Meta Spectrum Overview

First, let's examine the overall meta spectrum file that contains the full range of simulated spectra:

```{r meta-spectrum-plot, fig.cap="Sample of skin reflectance spectra colored by melanin index", fig.height=5}
# Generate plot from the meta spectrum file
meta_plot <- generateSkinSpectraPlotWrapper("../data/meta_spectrum_smooth.csv")
print(meta_plot)
```

# Comparison of Three Datasets

## Summary of Dataset Differences

**Only TWO parameters vary between datasets:**
1. **Thickness (TH)**: Skin thickness affecting light scattering
2. **Eumelanin (EU)**: Ratio of brown/black vs red/yellow pigment

All other parameters (Melanin, Blood Volume, Oxygenation) remain constant across all three datasets.

**"Narrow" refers to the parameter RANGES being narrower, not the skin itself.**

## Dataset 1: Training Data with Wide Parameter Ranges
- **Purpose**: Train models on the full range of skin thickness and eumelanin values
- **Thickness**: 0.000-1.000 (full range from very thin to thick skin)
- **Eumelanin**: 0.650-0.800 (moderate to high brown/black pigment)

## Dataset 2: Test Data with Wide Parameter Ranges  
- **Purpose**: Test model performance on same parameter space as training
- **Thickness**: 0.000-1.000 (matches training data)
- **Eumelanin**: 0.650-0.800 (matches training data)

## Dataset 3: Test Data with Narrow Parameter Ranges
- **Purpose**: Evaluate model on a specific subset - very thin skin with high eumelanin
- **Thickness**: 0.001-0.035 (only very thin skin - ~3.5% of the full range)
- **Eumelanin**: 0.700-0.780 (high eumelanin only - ~53% of the training range)
- **"Narrow" refers to**: Both thickness AND eumelanin parameter RANGES being narrower/restricted compared to training data
- **Physical interpretation**: This represents very thin skin (like eyelids) with predominantly brown/black pigmentation

## Side-by-Side Comparison of All Three Datasets

```{r three-plots-comparison, fig.width=15, fig.height=5}
# Create all three plots
folder_path <- "../data"
plots <- createThreeDatasetPlots(folder_path)

# Combine plots side by side using patchwork
if (length(plots) == 3) {
  combined_plot <- plots$train | plots$test_wide | plots$test_narrow
  combined_plot <- combined_plot + 
    plot_annotation(
      title = "Comparison of Training and Test Datasets",
      subtitle = "Each plot shows 400 randomly sampled spectra colored by melanin index"
    )
  print(combined_plot)
} else {
  cat("Note: Could not find all three dataset files.\n")
  # Display whatever plots were found
  for (name in names(plots)) {
    print(plots[[name]])
  }
}
```

### Complete Parameter Values for Each Dataset

```{r parameter-table, results='asis'}
# Create a table showing all parameters for each dataset.
param_table <- tibble::tibble(
  Parameter = c("Melanin (ME)", "Blood Volume (BL)", "<strong>Thickness (TH)</strong>",
                "<strong>Eumelanin (EU)</strong>", "Oxygenation (OX)"),
  `Training Dataset` = c("0.001-0.430", "0.000-0.200", "<strong>0.000-1.000</strong>",
                         "<strong>0.650-0.800</strong>", "0.900-1.000"),
  `Test Dataset (Wide)` = c("0.001-0.430", "0.000-0.200", "<strong>0.000-1.000</strong>",
                            "<strong>0.650-0.800</strong>", "0.900-1.000"),
  `Test Dataset (Narrow)` = c("0.001-0.430", "0.000-0.200", "<strong>0.001-0.035</strong>",
                              "<strong>0.700-0.780</strong>", "0.900-1.000")
)

knitr::kable(
  param_table,
  caption = "Parameter ranges for all three datasets. Bold values indicate parameters that vary between datasets.",
  align = c("l", "c", "c", "c"),
  escape = FALSE
)

cat(
  "\n::: {.callout-note}\n",
  "## Key differences\n",
  "- **Training & Test (Wide):** Identical parameters across the full physiological range.\n",
  "- **Test (Narrow):** Restricted ranges for both thickness (0.001-0.035) and eumelanin (0.700-0.780).\n",
  "- The narrow dataset tests whether models trained on broad ranges can generalize to a constrained subspace.\n",
  ":::\n",
  sep = ""
)

dataset_summary <- tibble::tibble(
  Dataset = c("Training", "Test (Wide)", "Test (Narrow)"),
  `ME (Melanin)` = c("0.001-0.430", "0.001-0.430", "0.001-0.430"),
  `BL (Blood)` = c("0.000-0.200", "0.000-0.200", "0.000-0.200"),
  `TH (Thickness)` = c("0.000-1.000", "0.000-1.000", "0.001-0.035"),
  `EU (Eumelanin)` = c("0.650-0.800", "0.650-0.800", "0.700-0.780"),
  `OX (Oxygenation)` = c("0.900-1.000", "0.900-1.000", "0.900-1.000")
)

knitr::kable(
  dataset_summary,
  caption = "Per-dataset parameter summary.",
  align = c("l", "c", "c", "c", "c", "c")
)

range_compare <- tibble::tibble(
  Parameter = c("Thickness (TH)", "Eumelanin (EU)"),
  `Training/Wide Range` = c("1.000", "0.150"),
  `Narrow Range` = c("0.034", "0.080"),
  `Narrow as % of Wide` = c("3.4%", "53.3%")
)

knitr::kable(
  range_compare,
  caption = "Range compression in the narrow test dataset.",
  align = c("l", "c", "c", "c")
)

cat(
  "\n```text\n",
  "THICKNESS (TH)\n",
  "Training/Wide: |==================================| 0.000 to 1.000\n",
  "Narrow:        |=|                                  0.001 to 0.035\n",
  "\n",
  "EUMELANIN (EU)\n",
  "Training/Wide: |==================================| 0.650 to 0.800\n",
  "Narrow:        |       ================|            0.700 to 0.780\n",
  "```\n",
  sep = ""
)
```

## Key Observations

### Training vs Test (Wide) Datasets
- Both contain the same parameter ranges
- This allows for direct validation of model performance
- Any differences in model performance can be attributed to the specific spectra sampled, not parameter range differences

### Test (Narrow) Dataset Characteristics  
- **"Narrow" means BOTH parameters are restricted:**
  - Thickness reduced from 0.000-1.000 to 0.001-0.035 (only ~3.5% of original range)
  - Eumelanin reduced from 0.650-0.800 to 0.700-0.780 (only ~53% of original range)
- Focuses exclusively on very thin skin with high eumelanin content
- Tests whether models trained on wide ranges can accurately predict spectra for this specific physiological subset
- Represents a challenging test case as it uses only a small portion of the training parameter space

## Implications for Model Development

1. **Generalization Testing**: The wide test set validates general model performance across the full parameter space
2. **Specific Case Testing**: The narrow test set evaluates performance on a challenging subset (very thin, high-eumelanin skin)
3. **Parameter Coverage**: 
   - Training covers the full physiological range
   - Test (Narrow) uses only ~3.5% of the thickness range and ~53% of the eumelanin range
   - This tests model interpolation capability within a restricted parameter space
